---
title: "Prediction of how well exercise is performed"
author: "demetr"
date: "July 17, 2016"
output: html_document
---

# Synopsis

In this report we analyze "Weight Lifting Exercises Dataset" dataset in order to form a prediction model which will be able to classify exercise to how well it was performed based on collected metrics.

# Reproducability

All calculations and modifying of data were performed by R and can be reproduced with this report. It assumes that  dataset was divided into two datasets (training and test) named `pml-training.csv` and `pml-training.csv`.

Following command is used to setup environment:

```{r , message = F}
library(caret)
library(dplyr)
library(xtable)

x <- Sys.setlocale("LC_TIME", "C")

set.seed(131313)
```


# Data set

We use dataset from paper [Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/har#dataset). This dataset was divided into two parts by "Practical Machine Learning" course teachers.

```{r , cache = TRUE}
exerciseTrain <- tbl_df(read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"))
exerciseTest <- tbl_df(read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"))
```

Both datasets contains data about numeric metrics collected during exercise, timestamps, names of persons and other information. Also dataset named "training" has variable "classe" which shows desired classification.

Training dataset has `r dim(exerciseTrain)[1]` observations while test dataset has only `r dim(exerciseTest)[1]`. Also training dataset has no data for several variables in most observed cases, e.g. variable `var_total_accel_belt` has `r sum(is.na(exerciseTrain$var_total_accel_belt))` empty observations.

It appears that dataset contains several timeseries observations, which is marked by `num_window` variable and belt metrics are collected only at the end of this interval.

## Data processing

We decide to modify original dataset in following way:

* Remove all timestamp and window information and treat all observation as independent.
* Remove all variables that are set only for end period.
* Remove information about who performed exercise.
* Divide training dataset into 3 dataset randomly: train dataset (60%), cross validation dataset (20%) and test dataset (20%).

```{r , cache = TRUE}
mapping <- sample(c(1, 2, 3), dim(exerciseTrain)[1], replace = T, prob = c(0.6, 0.2, 0.2))

exerciseTrain <- exerciseTrain %>% dplyr::select(
    roll_belt:total_accel_belt,
    gyros_belt_x:total_accel_arm,
    gyros_arm_x:magnet_arm_z,
    roll_dumbbell:yaw_dumbbell,
    gyros_dumbbell_x:yaw_forearm,
    gyros_forearm_x:magnet_forearm_z,
    classe)

trainDS <- exerciseTrain %>%
  filter(mapping[row_number()] == 1)
validationDS <- exerciseTrain %>%
  filter(mapping[row_number()] == 2)
testDS <- exerciseTrain %>%
  filter(mapping[row_number()] == 3)
```

# Models

We fit machine learning models agains all remaining variables. We fit random forest (“rf”) and linear discriminant analysis (“lda”) models.

```{r, cache = TRUE}
fitRf <- train(classe ~ ., data = trainDS, method = "rf", ntree = 10)
fitLda <- train(classe ~ ., data = trainDS, method = "lda")
```

To check effective accuracy we use validation dataset:

```{r, message = F, results="asis"}
validationTable <- rbind(
  confusionMatrix(predict(fitRf, validationDS), validationDS$classe)$overall, 
  confusionMatrix(predict(fitLda, validationDS), validationDS$classe)$overall)

rownames(validationTable) <- c("Random forest",
                               "Linear discriminant analysis")

print(xtable(validationTable), type = "html")
```

Based on validation results we fit combined model:

```{r, cache = TRUE}
combinedModelMatrix <- data.frame(
  predict(fitRf, validationDS),
  predict(fitLda, validationDS),
  validationDS$classe
)
colnames(combinedModelMatrix) <- c("rf", "lda", "classe")
fitComb <- train(classe ~ ., data = combinedModelMatrix)
```

We examine resulted accuracy on test dataset:

```{r, cache = TRUE}
combinedPredictMatrix <- data.frame(
  predict(fitRf, testDS),
  predict(fitLda, testDS)
)
colnames(combinedPredictMatrix) <- c("rf", "lda")

combinedPredict <- predict(fitComb, combinedPredictMatrix)
```

Resulted estimated accuracy for this prediction is:
```{r}
confusionMatrix(combinedPredict, testDS$classe)
```

## Estimation

We try to predict manner of exercise in `exerciseTest` dataset.

```{r}
combinedPredictMatrixTest <- data.frame(
  predict(fitRf, exerciseTest),
  predict(fitLda, exerciseTest)
)
colnames(combinedPredictMatrixTest) <- c("rf", "lda")

combinedPredictTest <- predict(fitComb, combinedPredictMatrixTest)

print(combinedPredictTest)
```